### 2014-12-16 doc/thesis/system.tex
Adding a section about interfaces and type aliases to chapter 3
Splitting the section about types and subtyping into two sections
Improvements to the section about types

\section{Types}

@@

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
t ::= & \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; p \rightarrow r & \textit{function types}\\
& | \; \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
l ::= & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
b ::= & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
v ::= & \;\; t \; | \; \Const \; t & \textit{value types}\\ 
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
s ::= & \;\; t &\\
& | \; t* & \textit{variadic types}\\
& | \; t \times s & \textit{tuple types}\\
p ::= & \;\; \Void \; | \; s & \textit{parameters list type}\\
r ::= & \;\; \Void \; | \; s \; | \; s \sqcup r & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

Figure \ref{fig:typelang} presents the abstract syntax of the
Typed Lua types.
Typed Lua splits types into two categories:
\emph{first-level types} and \emph{second-level types}.
Typed Lua uses first-level types to represent first-class Lua values and
second-level types to represent tuples of values that appear in 
multiple assignments and function applications.
First-level types include literal types, base types, the type $\Nil$,
the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
union types, function types, table types, recursive types, and
projection types.
Second-level types include the type $\Void$, tuples of first-level
types optionally ending in a variadic type, and unions of these tuples.

Literal types represent the type of literal values.
They can be the boolean values $\False$ and $\True$,
any integer value, any floating point value, and any string value.
We shall see that literal types are important in our treatment of
table types as records.

Our type system also includes the following base types: $\Boolean$,
$\Integer$, $\Number$, and $\String$.
The base types $\Boolean$ and $\String$ represent the values that during
run-time Lua tags as \texttt{boolean} and \texttt{string}, respectively.
Lua 5.3 introduced two internal representations to the tag \texttt{number}:
\texttt{integer} for integer numbers and \texttt{float} for real numbers.
Both \texttt{integer} and \texttt{float} values are subtype of \texttt{number}.
We introduced the base type $\Number$ to represent either the tag
\texttt{number} or \texttt{float} values, and the type $\Integer$ to represent
\texttt{integer} values.

@@

The type $\Nil$ represents the type of the value that during run-time
Lua tags as \texttt{nil}.
In our type system $\Nil$ is not the bottom type, and it is subtype
only of itself through the reflexive rule.
Typed Lua has this behavior because in Lua the tag \texttt{nil}
represents the absence of an useful value.
Thus, making $\Nil$ the bottom type would lead to several expressions
that would pass the type checker, but that would fail during run-time
in the presence of a \texttt{nil} value.

The type $\Value$ is the top type, so any first-level type is a
subtype of $\Value$.
In the next section we will show that this type along with
variadic types help the type system drop extra values on
multiple assignments and function calls, thus preserving the
semantics of Lua in these cases.
Following we show its subtyping rule:
\[
\begin{array}{c}
\mylabel{S-VALUE}\\
\senv \vdash t \subtype \Value
\end{array}
\]

The dynamic type $\Any$ allows programmers to mix static and dynamic typing;
it is neither the bottom nor the top type,
but a separate type that is subtype only of itself through the reflexive rule.

@@

Typed Lua uses the type $\Self$ to represent the \emph{receiver}
in object-oriented method definitions and method calls.
As we mentioned in Chapter \ref{chap:typedlua}, we need the type
$\Self$ to prevent programmers from indexing a method without
calling it with the correct \emph{receiver}.
The type $\Self$ is subtype only of itself through the reflexive rule.

Union types represent types that can hold a value of several
different, but fixed types.
The subtyping rules for union types are standard:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNION1}\\
\dfrac{\senv \vdash t_{1} \subtype t \;\;\;
       \senv \vdash t_{2} \subtype t}
      {\senv \vdash t_{1} \cup t_{2} \subtype t}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION2}\\
\dfrac{\senv \vdash t \subtype t_{1}}
      {\senv \vdash t \subtype t_{1} \cup t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION3}\\
\dfrac{\senv \vdash t \subtype t_{2}}
      {\senv \vdash t \subtype t_{1} \cup t_{2}}
\end{array}
\end{array}
\]

@@

To introduce function types we first need to introduce
second-level types, as they appear on both sides of function types.
Second-level types are tuples of first-level types
that can optionally end with a variadic type.
Our type system needs second-level types because tuples are
not first-class values in Lua, though they may appear on
argument passing, multiple returns, and multiple assignments.
A variadic type $t*$ is a generator for a sequence of values of
type $t \cup \Nil$, and it represents the type of the vararg
expression.
The subtyping rule for tuple types is standard, but the subtyping
rules for variadic types are not obvious:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TUPLE}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash s_{1} \subtype s_{2}}
      {\senv \vdash t_{1} \times s_{1} \subtype t_{2} \times s_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VARARG1}\\
\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
      {\senv \vdash t_{1}{*} \subtype t_{2}{*}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG2}\\
\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2}}
      {\senv \vdash t_{1}{*} \subtype t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG3}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil}
      {\senv \vdash t_{1} \subtype t_{2}{*}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VARARG4}\\
\dfrac{\senv \vdash t_{1}{*} \subtype t_{2} \;\;\;
       \senv \vdash t_{1}{*} \subtype s_{2}}
      {\senv \vdash t_{1}{*} \subtype t_{2} \times s_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG5}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2}{*} \;\;\;
       \senv \vdash s_{1} \subtype t_{2}{*}}
      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}{*}}
\end{array}
\end{array}
\]

@@

Function types have the form $p \rightarrow r$, where
$p$ is the type of the parameter list and $r$ is the return type.
The type of the parameter list can be $\Void$ or a tuple type,
while the return type can be $\Void$, a tuple type, or a union of
tuple types.
The type $\Void$ means an empty tuple.
Our type system includes unions of tuples on the return type because
in Lua it is common to overload the return type to denote error,
as we mentioned in Section \ref{sec:statistics}.

The type $\Void$ and union of tuple types are also second-level types.
The type $\Void$ is subtype only of itself through the reflexive rule.
The subtyping rules for unions of tuple types are similar to the
subtyping rules for unions of first-level types.

@@

Table types represent the various forms that Lua tables can take.
The syntactical form of table types is $\{ k_{i}{:}v_{i}, ..., k_{n}{:}v_{n} \}_{tag}$.
When $n = 0$, then $i = 0$ and it represents the type of an empty table.
When $n > 0$, then $i$ varies from 1 to $n$, each $k_{i}$ represents
the type of a table key, and each $v_{i}$ represents the type of the
value that table keys of type $k_{i}$ map to.
Key types can only be literal types, base types, and the dynamic type.
We made this restriction to the type of the keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The dynamic type is an option when we need a loose table type.
Value types can be any first-level type, and can optionally include
the $\Const$ attribute to denote immutable values.

We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
to classify table types.
The first tag represents the table types that the type annotations,
interface, and userdata declarations describe.
The second tag represents the table types that we can use field assignment
to add new fields to them.
The third tag represents the table types of table constructors.
We have different subtyping rules for table types according to their
tags.

@@

Recursive types have the form $\mu x.t$,
where $t$ is a first-level type with $x$ appearing anywhere this
first-level type could appear.
For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
is a type for singly-linked list of integers.

@@

Projection types have the form $\pi_{i}^{x}$, where $x$ is a variable
that stores an union of tuple types in a special type environment,
and $i$ is an index that indicates how to build a first-level
union type from $x$.
For instance, if $x$ is the union of tuple types
$(\Integer \times \Integer) \sqcup (\Nil \times \String)$,
then $\pi_{1}^{x}$ represents the type $\Integer \cup \Nil$
while $\pi_{2}^{x}$ represents the type $\Integer \cup \String$.
Projection types follow the reflexive subtyping rule.
In the next section we will show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear on the right-hand side of the declaration of local variables.
We will also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

\textit{/* How should we control invalid indices? */}

\section{Typing rules}

>>>>

\section{Types}
\label{sec:types}

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
t ::= & \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; s \rightarrow r & \textit{function types}\\
& | \; \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
l ::= & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
b ::= & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
v ::= & \;\; t \; | \; \Const \; t & \textit{value types}\\ 
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
s ::= & \;\; \Void & \textit{void type}\\
& | \; t* & \textit{variadic types}\\
& | \; t \times s & \textit{tuple types}\\
r ::= & \;\; s \; | \; s \sqcup r & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

Figure \ref{fig:typelang} presents the abstract syntax of the
Typed Lua types.
Typed Lua splits types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear in 
assignments and function applications.
First-level types include literal types, base types, the type $\Nil$,
the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
union types, function types, table types, recursive types, and
projection types.
Second-level types include the type $\Void$, variadic types,
tuple types, and unions of tuple types.

Literal types $l$ represent the type of literal values.
They can be the boolean values $\False$ and $\True$,
an integer value, a floating point value, or a string value.
We shall see that literal types are important in our treatment of
table types as records.

Typed Lua includes four base types $b$: $\Boolean$, $\Integer$, $\Number$, and $\String$.
The base types $\Boolean$ and $\String$ represent the values that during
run-time Lua tags as \texttt{boolean} and \texttt{string}, respectively.
Lua 5.3 introduced two internal representations to the tag \texttt{number}:
\texttt{integer} for integer numbers and \texttt{float} for real numbers.
Both \texttt{integer} and \texttt{float} values are a subtype of \texttt{number}.
We introduced the base type $\Number$ to represent either the tag
\texttt{number} or \texttt{float} values, and the type $\Integer$ to represent
\texttt{integer} values.
In the next section we will show that $\Integer$ is subtype of $\Number$.
This allows programmers to use \texttt{integer} values in the
place of \texttt{float} values.

The type $\Nil$ represents the type of the value that during run-time
Lua tags as \texttt{nil}.

The type $\Value$ is the top type.
In Section \ref{sec:rules} we will show that this type,
along with variadic types, helps the type system to drop extra values
on assignments and function calls, thus preserving the
semantics of Lua in these cases.

Typed Lua includes the dynamic type $\Any$ for allowing programmers
to mix static and dynamic typing.

Typed Lua uses the type $\Self$ to represent the \emph{receiver}
in object-oriented method definitions and method calls.
As we mentioned in Chapter \ref{chap:typedlua}, we need the type
$\Self$ to prevent programs from indexing a method without
calling it with the correct receiver.

Union types $t_{1} \cup t_{2}$ represent types that can hold a value
of two different types.

Function types have the form $s \rightarrow r$ to represent Lua functions,
where $s$ is the type of the parameter list and $r$ is the return type.
Both $s$ and $r$ are second-level types.
In the next paragraphs we explain why we need different representations
for the type of the parameter list and the return type.

Second-level types $s$ are tuples of first-level types that can end
with either an empty tuple or with a variadic type.
Typed Lua needs second-level types because tuples are not first-class
values in Lua, though they may appear on argument passing,
multiple returns, and multiple assignments.
The type $\Void$ is the type of an empty tuple.
A variadic type $t*$ is a generator for a sequence of values
of type $t \cup \Nil$, and it represents the type of a vararg expression.

Second-level types $r$ include the same second-level types introduced
by $s$ plus unions of tuple types.
Typed Lua includes unions of tuples because Lua programs
usually overload the return type of functions to denote error,
as we mentioned in Section \ref{sec:statistics}.
We could not find a nice concrete syntax to represent unions of
tuples that appear in the type of the parameter list, and
that is the reason why we restrict them to appear in the return type only.
The same reason led us to split second-level types into two
different symbols.
We use the symbol $\sqcup$ to represent the union between two
different tuple types.
We have different unions because $\cup$ represents the union
between two first-level types, while $\sqcup$ represents the
union between two tuple types.

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{tag}$,
where each $k_{i}$ represents the type of a table key,
and each $v_{i}$ represents the type of the value that table keys of type $k_{i}$ map to.
Key types can only be literal types, base types, or the dynamic type.
We made this restriction to the type of the keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The dynamic type is an option when we need a loose table type.
Value types can be any first-level type, and can optionally include
the $\Const$ type to denote immutable values.

We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
to classify table types.
The tag \emph{unique} represents tables that we know their contents,
and that there is no alias to them.
In particular, the type of the table constructor has this tag.
The tag \emph{open} represents \emph{unique} table types that
have at least one alias.
The tag \emph{closed} represents table types that do not provide
any guarantees about the types not listed in the table type.
In particular, type annotations as well as interfaces and userdata
declarations always describe \emph{closed} table types.
In the next sections we explain in more detail why we need
different table types.

Recursive types have the form $\mu x.t$,
where $t$ is a first-level type that $x$ represents.
For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
is a type for singly-linked lists of integers.

Type Lua includes projection types as a way to project
unions of tuple types into unions of first-level types.
Projection types have the form $\pi_{i}^{x}$, where $x$ is a variable
that stores an union of tuple types in a special type environment,
and $i$ is an index that indicates how to project a first-level
union type from $x$.
For instance, if $x$ is the union of tuple types
$(\Integer \times \Integer) \sqcup (\Nil \times \String)$,
then $\pi_{1}^{x}$ represents the type $\Integer \cup \Nil$
while $\pi_{2}^{x}$ represents the type $\Integer \cup \String$.
In Section \ref{sec:rules} we show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear on the right-hand side of the declaration of local variables.
We also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

\section{Subtyping}
\label{sec:subtyping}

@@

@@

In our type system $\Nil$ is not the bottom type, and it is subtype
only of itself through the reflexive rule.
Typed Lua has this behavior because in Lua the tag \texttt{nil}
represents the absence of an useful value.
Thus, making $\Nil$ the bottom type would lead to several expressions
that would pass the type checker, but that would fail during run-time
in the presence of a \texttt{nil} value.

Any first-level type is a subtype of $\Value$:
\[
\begin{array}{c}
\mylabel{S-VALUE}\\
\senv \vdash t \subtype \Value
\end{array}
\]

The dynamic type $\Any$ is neither the bottom nor the top type,
but a separate type that is subtype only of itself through the reflexive rule.

@@

The type $\Self$ is subtype only of itself through the reflexive rule.

The subtyping rules for union types are standard:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNION1}\\
\dfrac{\senv \vdash t_{1} \subtype t \;\;\;
       \senv \vdash t_{2} \subtype t}
      {\senv \vdash t_{1} \cup t_{2} \subtype t}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION2}\\
\dfrac{\senv \vdash t \subtype t_{1}}
      {\senv \vdash t \subtype t_{1} \cup t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION3}\\
\dfrac{\senv \vdash t \subtype t_{2}}
      {\senv \vdash t \subtype t_{1} \cup t_{2}}
\end{array}
\end{array}
\]

@@

The subtyping rule for tuple types is standard, but the subtyping
rules for variadic types are not obvious:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TUPLE}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash s_{1} \subtype s_{2}}
      {\senv \vdash t_{1} \times s_{1} \subtype t_{2} \times s_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VARARG1}\\
\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
      {\senv \vdash t_{1}{*} \subtype t_{2}{*}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG2}\\
\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2}}
      {\senv \vdash t_{1}{*} \subtype t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG3}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil}
      {\senv \vdash t_{1} \subtype t_{2}{*}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VARARG4}\\
\dfrac{\senv \vdash t_{1}{*} \subtype t_{2} \;\;\;
       \senv \vdash t_{1}{*} \subtype s_{2}}
      {\senv \vdash t_{1}{*} \subtype t_{2} \times s_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG5}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2}{*} \;\;\;
       \senv \vdash s_{1} \subtype t_{2}{*}}
      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}{*}}
\end{array}
\end{array}
\]

@@

The type $\Void$ is subtype only of itself through the reflexive rule.
The subtyping rules for unions of tuple types are similar to the
subtyping rules for unions of first-level types.

@@

@@

@@

Projection types follow the reflexive subtyping rule.

\section{Typing rules}
\label{sec:rules}
